A formal approach but practical approach in train regulation

### 2025-06-26 paper/main.tex
cas des aiguilles dans la composition

\begin{document}
%
\title{Formal verification of train route planning}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Lucas Villaume \and Guillaume Bonfante \and Martin Vassor}
%
\authorrunning{L. Villaume et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Université de Lorraine, CNRS, Inria, LORIA, F-54000 Nancy, France}
%
\maketitle
%
\begin{abstract}
	% Contexte
	Trains networks are critical system which failure may lead to numerous fatalities, in case of crashes, or important loss, if trains are not properly routed. 
	% Problème
	In this paper, our goal is to formally verify the correctness of train routes. More precisely, we verify the absence of deadlocks, the termination, and the absence of crashes.
	% Approche
	To address this question, we model train systems on paper and in TLA+. Our model includes trains, as well as switches, detection sensors and traffic lights. Both switches and traffic lights are controlled by a regulator upon train events. We provide trains and the regulator with orders, characterising the schedule, which is then checked using TLA+'s model checker.
	% Resultats
	Our contributions are threefold: first, we provide formal model of railroad networks and schedules; second, we implement this model as a TLA+ specification and we report tests on checking schedules on a non-trivial network, composed of 3 trains, 8 blocks and 5 switches; finally we show how to sequentially compose schedules, which we check against our TLA+ specification.

@@

Over the last decades, we observe a general and increasing interest of formal methods for verification and validation of safety-critical systems, including railway networks.
In \cite{Ferrari_Formal_2022}, Ferrari et al. present a survey of use cases of formal methods in railways development. They identify various phases of development phases where formal methods have been used, ranging from planning to maintenance via implementation, testing, etc., as well as most frequently studied subsystems (interlocking systems being the most formally verified component). In this paper, focus specifically on train \emph{routing}, i.e. finding correct train schedules (for some correctness criteria). This topic is related to that of interlocking (we prevent trains from using simultaneously some elements), but from the viewpoint of trains and their routes, while the viewpoint of interlocking is mostly trackworks.

In~\cite{lusby_railway_2011}, Lusby et al.\ survey works on train routing. In this context, we can distinguish two problems: automated routing and routing verification. The former deals with automatically finding train routes satisfying some properties. Lusby et al.\ mention that solving this problem, in highly interconnected network, is not feasible in practice. In this paper, we address the second problem: verifying whether a train routing satisfies correctness criteria (without considering how it is obtained).

@@

Essentially, this guard is the state machine shown in Figure~\ref{fig:state_machine_guard}. The four states of the guard are \guardR, \guardS, and \guardT.

\begin{figure}
	\centering
	\begin{tikzpicture}
		\node[draw] (trains) {Train movements (\guardT)};
		\node[left=.5cm of trains.west] (zero) {};
		\node[draw, right=of trains] (regulator) {Regulator reaction (\guardR)};
		\node[draw, below=of regulator] (signals) {Signals update (\guardS)};

		\draw[-stealth] (zero) to (trains);
		\draw[-stealth] (trains) to[bend right=10] (signals);
		\draw[-stealth] (signals) to[bend right=10] (trains);
		\draw[-stealth] (trains) to (regulator);
		\draw[-stealth] (regulator) to (signals);
	\end{tikzpicture}
	\caption{State machine of the guard.}
	\label{fig:state_machine_guard}
\end{figure}

@@

States of the system are 4-tuples from $\{\guardR, \guardS, \guardT\} \times \bufTrainSet \times \bufSigSet \times \modelSet$. Our reduction relation is a binary relation on system states, noted $\reduces$: $S_1 \reduces S_2$ means that state $S_1$ can change to $S_2$ (note that this relation is not a function: $S_1$ can reduce to multiple states, due to non-determinism). We define $\reduces$ as the least relation induced by the inference rules presented hereafter.

@@

When the regulator reacts to a train event, it executes atomically all rules associated with that event. To implement this atomic execution, the first regulator event is to change the guard to \guardR{} (see Fig. \ref{fig:state_machine_guard}).

@@

Rule \ruleRef{Turn} triggers when a \turn{\sid{id}}{s} is encountered in the event handler. It updates the state of the targeted switch.
\begin{mathpar}
	\inferrule*[left=\ruleDef{Turn}]{
		\bufhead{\bufTrain} = \tid{id}
		\and
		\head{\handlerOf{E}{\tid{id}}} = \handler{\bid{id}}{\regConcat{\turn{\sid{id}}{s}}{A^\prime}}
	}{
	\redTuple{\guardR}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regTuple{E}{T}{W}}{\switches}{\signals}}
	\\\reduces
	\redTuple{\guardR}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regTuple{\nextAct{\handlerOf{E}{\tid{id}}}}{T}{W}}{\updateSwitch{\switches}{\sid{id}}{s}}{\signals}}
	}
\end{mathpar}

@@

To each safe step $X = (\vec{x}^{\tid{i}})_{\tid{i} \in \trains}$ coming with a network configuration $\sigma$ and a direction map $d$, just below, we define the system state $S(X) = \redTuple{\guardI}{ \emptyList}{ \emptyList}{\stateTuple{\trainSeq{}}{\regulator}{\switches}{\signals}}$, next called a \emph{safe step state}. Informally, we set turnouts to their position according to $\sigma$ during the initial step  and we lock signals to red for each train's destination according to $d$. 
\begin{itemize}
\item for each train $\tid{i}$, $\trainSeq[\tid{i}] = \trainTuple{\tid{i}}{x_1^{\tid{i}}}{\dirStop}{\su{d(\tid{i})}{\vec{x}^{\tid{i}}}}$,
\item $R = \langle E,  \bid{i} \in \blocks \mapsto 0, W\rangle$, $E$ being defined just below,
\item $W = \{ \tuple{x_1^\tid{i}, x_1^\tid{i}, 1, \tid{i}}  \mid \tid{i} \in \trains \}$, 
\item $F = \blocks \times \directions \to \{\sigred, \siggreen \}$: 
$$\langle \bid{i}, d\rangle \mapsto \left \{ \begin{array}{ll}
\sigred & \text{ if } \bid{i} = x_1^\tid{j}  \text{ for some }j \in \trains\\%  \\
\siggreen & \text{otherwise}
\end{array}\right.
$$
\end{itemize}

@@

So, we suppose we are given safe step states $S_1, \ldots, S_k$ corresponding to $X_1, \ldots, X_k$. To compute their composition $S_1 \cdot S_2 \cdots S_k$, we proceed by induction on $k$. For $k = 1$, we return $S_1$. For $k>0$, let $S = S_1 \cdots S_{k-1}$. Let $S = \langle \guardG, \bufTrain, \bufSig, \langle \Gamma, \regulator, \switches, \signals\rangle \rangle$. We set $S \cdot S_k = \langle \guardG, \bufTrain, \bufSig, \langle \Gamma', \regulator', \switches, \signals\rangle \rangle$ as follows. First,  for any train $\tid{i} \in \trains$, let $P_0 = S.\Gamma[\tid{i}].P$ and $P_k = S_k.\Gamma[\tid{i}].P$, then set $\Gamma'[\tid{i}] = \Gamma[\tid{i}][ P \leftarrow P_0 \cdot P_k]$.

Second, to define $E'$, we start $E' = E$ and we enrich it as follows.  For each

@@

Currently, our model does not include timing aspects. While we can characterise correctness properties, we lack the ability to specify \emph{quality of service} constraints. We intend to add time to our model, in order to be able to specify how fast trains ought to reach their destinations.

Another aspect we did not explore is reordering trains. So far, train schedules are decided statically, which may lead to unwanted situations: e.g. a delayed train supposed to use a block first prevents other trains to use that block during the delay. Dynamic reordering of trains, while difficult to implement, could improve this situation. To a lesser extend, we could extend our model with alternatives: having a static schedule that allow multiple possible ordering, and deciding which is taken at runtime.

@@

Experiments presented in this paper were carried out using the Grid'5000 testbed, supported by a scientific interest group hosted by Inria and including CNRS, RENATER and several Universities as well as other organizations (see \url{https://www.grid5000.fr}).

>>>>

\begin{document}
%
\title{A formal approach but practical approach in train regulation}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Lucas Villaume \and Guillaume Bonfante \and Martin Vassor}
%
\authorrunning{L. Villaume et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Université de Lorraine, CNRS, Inria, LORIA, F-54000 Nancy, France}
%
\maketitle
%
\begin{abstract}
	% Contexte
	Trains networks are critical system which failure may lead to numerous fatalities, in case of crashes, or important loss, if trains are not properly routed. 
	% Problème
	In this paper, our goal is to formally verify the correctness of train routes. More precisely, we verify the absence of deadlocks, the termination, and the absence of crashes.
	% Approche
	To address this question, we model train systems on paper and in TLA+. Our model includes trains, as well as switches, detection sensors and traffic lights. Both switches and traffic lights are controlled by a regulator upon train events. We provide trains and the regulator with orders, characterising the schedule, which is then checked using TLA+'s model checker.
	% Resultats
	Our contributions are threefold: first, we provide formal model of railroad networks and schedules; second, we implement this model as a TLA+ specification and we report tests on checking schedules on a non-trivial network, composed of 3 trains, 8 blocks and 5 switches; finally we show how to sequentially compose schedules, which we check against our TLA+ specification.

@@

@@

Essentially, this guard is the state machine shown in Figure~\ref{fig:state_machine_guard}. The four states of the guard are \guardI, \guardR, \guardS, and \guardT.

\begin{figure}
	\centering
	\begin{tikzpicture}
		\node[] (zero) {};
		\node[draw, below=of zero] (initial) {Initial state (\guardI)};
		\node[draw, below=of initial] (trains) {Train movements (\guardT)};
		\node[draw, right=of initial] (signals) {Signals update (\guardS)};
		\node[draw, right=of trains] (regulator) {Regulator reaction (\guardR)};

		\draw[-stealth] (zero) to (initial);
		\draw[-stealth] (initial) to (signals);
		\draw[-stealth] (trains) to[bend right=10] (signals);
		\draw[-stealth] (signals) to[bend right=10] (trains.north);
		\draw[-stealth] (trains) to (regulator);
		\draw[-stealth] (regulator) to (signals);
	\end{tikzpicture}
	\caption{State machine of the guard.}
	\label{fig:state_machine_guard}
\end{figure}

@@

States of the system are 4-tuples from $\{\guardI, \guardR, \guardS, \guardT\} \times \bufTrainSet \times \bufSigSet \times \modelSet$. Our reduction relation is a binary relation on system states, noted $\reduces$: $S_1 \reduces S_2$ means that state $S_1$ can change to $S_2$ (note that this relation is not a function: $S_1$ can reduce to multiple states, due to non-determinism). We define $\reduces$ as the least relation induced by the inference rules presented hereafter.

\paragraph{Initialisation rules.}
Upon starting, we first set up the switches and the signals using rules \ruleRef{TurnInit} and \ruleRef{SigInit}.

\begin{mathpar}
	\inferrule*[left=\ruleDef{TurnInit}]{
		\head{\actionsOf{\head{\initialHandler{E}}}} = \turn{\sid{id}}{s}
	}{
		\redTuple{\guardI}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regTuple{E}{T}{W}}{\switches}{\signals}}
		\\\reduces
		\redTuple{\guardI}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regTuple{\nextAct{\initialHandler{E}}}{T}{W}}{\updateSwitch{\switches}{\sid{id}}{s}}{\signals}}
	}
\end{mathpar}

\begin{mathpar}
	\inferrule*[left=\ruleDef{SigInit}]{
		\head{\actionsOf{\head{\initialHandler{E}}}} = \sendred{\bid{id}}{\dirFmt{d}}
	}{
	\redTuple{\guardI}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regTuple{E}{T}{W}}{\switches}{\signals}}
	\\\reduces
		\redTuple{\guardI}{\bufTrain}{\push{\bufSig}{\tuple{\bid{id}, \sigred}}}{\stateTuple{\trainSeq}{\regTuple{\nextAct{\initialHandler{E}}}{T}{W}}{\switches}{\signals}}
	}
\end{mathpar}\todo{Est-ce qu'on a vraiment besoin de spécifier la direction ?}

Finally, when initial actions are done, we start the system by switching the guard to state \guardS. Switching to that state instead of \guardT ensures signals perform requests of the initialisation step, as well as an first computation of the safety rules (see \ruleRef{SignalReq} and \ruleRef{SignalSfty} below for signals dynamics).

\begin{mathpar}
	\inferrule*[left=\ruleDef{TurnInit}]{
	\actionsOf{\head{\initialHandler{E}}} = \emptyList
	}{
		\redTuple{\guardI}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regTuple{E}{T}{W}}{\switches}{\signals}}
		\reduces
		\redTuple{\guardS}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regTuple{E}{T}{W}}{\switches}{\signals}}
	}
\end{mathpar}

@@

When the regulator reacts to a train event, it executes atomically all rules associated with that event. To implement this atomic execution, the first regulator event is to change the guard to \guardR (see Fig. \ref{fig:state_machine_guard}).

@@

Rule \ruleRef{Turn} triggers when a \turn{\sid{id}}{s} is encountered in the event handler. It updates the state of the targeted switch.
\begin{mathpar}
	\inferrule*[left=\ruleDef{Turn}]{
		\bufhead{\bufTrain} = \tid{id}
		\and
		\head{\handlerOf{E}{\tid{id}}} = \handler{\bid{id}}{\regConcat{\turn{\sid{id}}{s}}{A^\prime}}
	}{
	\redTuple{\guardR}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regulator}{\switches}{\signals}}
	\\\reduces
	\redTuple{\guardR}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regulator[E \leftarrow E[ \tid{id} \leftarrow \langle \bid{id}, A'\rangle] ] }{\updateSwitch{\switches}{\sid{id}}{s}}{\signals}}
	}
\end{mathpar}
\todo{On ne peut pas résumer le régulateur à $R$, sinon $E$ n'est pas déclarée/n'apparait nulle part. Ou alors il faut préciser des notations comme $R.E$. On ne peut pas utiliser $R[E \leftarrow...]$ car $R$ n'est pas une fonction mais un tuple.}

@@

To each safe step $X = (\vec{x}^{\tid{i}})_{\tid{i} \in \trains}$ coming with a network configuration $\sigma$ and a direction map $d$, just below, we define the system state $S(X) = \redTuple{\guardI}{ \emptyList}{ \emptyList}{\stateTuple{\trainSeq{}}{\regulator}{\switches}{\signals}}$, next called a \emph{safe step state}. Informally, we set turnouts to their position according to $\sigma$ during the initial step  and we lock signals to red for each train's destination according to $d$. 
\begin{itemize}
\item for each train $\tid{i}$, $\trainSeq[\tid{i}] = \trainTuple{\tid{i}}{x_1^{\tid{i}}}{\dirStop}{\su{d(\tid{i})}{\vec{x}^{\tid{i}}}}$,
\item $R = \langle E,  \bid{i} \in \blocks \mapsto 0, W\rangle$, $E$ being defined just below,
\item $W = \{ \tuple{x_1^\tid{i}, x_1^\tid{i}, 1, \tid{i}}  \mid \tid{i} \in \trains \}$, 
\item $F = \blocks \times \directions \to \{\sigred, \siggreen \}$ defined for $\bid{i} \in \blocks, d \in \directions$ by: $F(\langle \bid{i}, d\rangle) = \sigred$ if  $\bid{i} = x_1^\tid{j}$ for some $j\in \trains$, otherwise $F(\langle \bid{i}, d\rangle)  = \siggreen$. 
\end{itemize}

@@

So, we suppose we are given safe step states $S_1, \ldots, S_k$ corresponding to $X_1, \ldots, X_k$. To compute their composition $S_1 \cdot S_2 \cdots S_k$, we proceed by induction on $k$. For $k = 1$, we return $S_1$. For $k>0$, let $S = S_1 \cdots S_{k-1}  = \langle \guardG, \bufTrain, \bufSig, \langle \Gamma, \langle E, W, T\rangle, \switches, \signals\rangle \rangle$. We set $S \cdot S_k = \langle \guardG, \bufTrain, \bufSig, \langle \Gamma', \langle E', W, T\rangle, \switches', \signals\rangle \rangle$ as follows.

First,  for any train $\tid{i} \in \trains$, let $P_0 = S.\Gamma[\tid{i}].P$ and $P_k = S_k.\Gamma[\tid{i}].P$, then set $\Gamma'[\tid{i}] = \Gamma[\tid{i}][ P \leftarrow P_0 \cdot P_k]$.

Second, we proceed by updates of $E$ and $\sigma$. At first, we set $E' = E$ and $\sigma' = \sigma$. Now, suppose that $\sigma_k$ is the network configuration coming with $X_k$. Let $\sigma_0 = \{ \sid{i} \mid \sigma(\sid{i}) \neq  \sigma_k(\sid{i})\}$. Given  $\sid{i} \in \turnouts$, suppose there is a step $j < k$, a train $\tid{m}$ such that $\sid{i} \in \throughS(x_n, d, \sigma_j, x_{n+1})$ with $\vec{x}$ being the route of train $\tid{m}$ at step $X_j$ and $\sigma_j$ the network configuration associated to $j$. Then, we keep the triple  $\langle j, \tid{m}, n\rangle$ with the highest value $j$. If there is no such triple, set $\sigma' := \sigma'[ \sid{i} \leftarrow \sigma_k(\sid{i})]$.  Otherwise, set $E' := E'[ ....]$.

@@

@@
